---
title: "Reproducible Research: Peer Assessment 1"
author: "Steven Fu"
date: "Saturday, November 15, 2014"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r readingdata}
setwd("C:\\Users\\I309666\\github\\RepData_PeerAssessment1")
unzip("activity.zip")
data<-read.csv("activity.csv")
data$date<-as.Date(data$date)
```

## What is mean total number of steps taken per day?
```{r plot1, echo=TRUE}
# to check if plyr package is installed, install if necessary
is.installed <- function(x) is.element(x, installed.packages()[,1]) 
if(!is.installed("plyr")){
        install.packages("plyr")
}
library(plyr)

stepByDay <- ddply(data, .(date), summarize, steps=sum(steps))
hist(stepByDay$steps, col="blue",
     main="Plot 1 - Histogram of number of steps per day",
     xlab="number of steps taken", ylab="frequency")
mean<-mean(stepByDay$steps, na.rm=TRUE)
mean
median<-median(stepByDay$steps, na.rm=TRUE)
median
```

####Average number of steps taken per day is `r as.integer(mean)`.
####Median of number of steps taken per day is `r as.integer(median)`.


## What is the average daily activity pattern?
```{r plot2, echo=TRUE}
stepByInterval <- ddply(data, .(interval), summarize, steps=mean(steps,na.rm = TRUE))
plot(x = stepByInterval$interval, y = stepByInterval$steps,type="l", col="blue",
     main="Plot 2 - Average number of steps across all interval",
     xlab="interval", ylab="average number of steps")
max<-stepByInterval[stepByInterval$steps==max(stepByInterval$steps),]
max
```
####The interval that contains on average the maximum number of steps 
####is `r max$steps` at interval `r max$interval`.


## Imputing missing values
```{r missingvalue, echo=TRUE}
missing<-is.na(data$steps)
sum_missing<-sum(missing)
```
The number of missing observations is `r sum_missing`.

```{r replacemissingvalue, echo=TRUE}
data2<-data
head(data2)
data2$intervalmean<-ave(data2$steps,data2$interval, FUN=function(x) mean(x[!is.na(x)]))
data2$steps[is.na(data2$steps)]<-data2$intervalmean[is.na(data2$steps)]
head(data2)
```

```{r plot3, echo=TRUE}
stepByDay2 <- ddply(data2, .(date), summarize, steps=sum(steps,na.rm = TRUE))
hist(stepByDay2$steps, col="blue",
     main="Plot 3 - Histogram of number of steps per day\nwith imputing missing values",
     xlab="number of steps taken", ylab="frequency")
mean2<-mean(stepByDay2$steps, na.rm=TRUE)
mean2
median2<-median(stepByDay2$steps, na.rm=TRUE)
median2
```

####Average number of steps taken per day is `r as.integer(mean2)`.
####Median of number of steps taken per day is `r as.integer(median2)`.
####These values are very close to the estimates from the first part of the assignment.


## Are there differences in activity patterns between weekdays and weekends?
```{r plot4, echo=TRUE}
weekday <- function(date) {
        day<- weekdays(date, abbreviate =TRUE)
    if (day %in% c("Mon", "Tue","Wed","Thu","Fri"))
		return("Weekday")
	else if (day %in% c("Sat","Sun"))
		return("Weekend")
	else
		stop("Invalid Date")
}
data2$weekday<-factor(sapply(data2$date, FUN=weekday))
head(data2)
stepByInterval2 <- ddply(data2, .(interval,weekday), summarize, steps=mean(steps,na.rm = TRUE))
head(stepByInterval2)

if(!is.installed("ggplot2")){
        install.packages("ggplot2")
}
library(ggplot2)
ggplot(data=stepByInterval2, aes(interval, steps)) + geom_line() +
        facet_grid(weekday ~ . ) + xlab("interval")+ ylab("average number of steps")+
        ggtitle("Plot 4 - Average number of steps across all interval\nweekday vs weekend")
```

####Total number of steps taken between interval  500 and 1000 is generally higher on weekdays.
####Total number of steps taken between interval 1000 and 2000 is generally higher on weekends.