---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


Reproducible Research: Peer Assessment 1  
==========================================
<br>


## Loading and preprocessing the data
Import the data
```{r}
activity <- read.table(file = unz("activity.zip", "activity.csv"), 
                       header = TRUE, 
                       sep = ",", 
                       colClasses = c("integer", 
                                      "Date", 
                                      "integer"))
```
Modify the data
```{r}
## use the dplyr package
suppressMessages(library(dplyr))

## create more important time based fields
activity <- activity %>% 
    mutate(datetime = as.POSIXct((floor(interval/100)*60 + 
                                      interval %% 100)*60, 
                                 origin = date, 
                                 tz = "UTC"), 
           minutes = floor(interval/100)*60 + interval %% 100, 
           time = format(datetime, "%H:%M")) %>% 
    arrange(date, interval)
```
<br>


## What is mean total number of steps taken per day?
First, sum the number of steps taken to the daily level
```{r}
daily <- activity %>% 
    group_by(date) %>% 
    summarise(numSteps = sum(steps, na.rm = TRUE)) %>% 
    arrange(date)
```
Next, create a histogram of the total number of steps taken each day
```{r, fig.width=10, fig.height=6}
## use the ggplo2 package (scales for axis text)
suppressMessages(library(ggplot2))
suppressMessages(library(scales))

## histogram
ggplot(data = daily, 
       aes(x = numSteps)) + 
    geom_histogram(col = "black", 
                   fill = "steelblue", 
                   binwidth = 1000) + 
    xlab("Total Number of Steps Taken Each Day") + 
    ylab("Number of Days") + 
    scale_x_continuous(limits = c(0, 25000), 
                       breaks = seq(0, 25000, 5000), 
                       labels = comma) + 
    scale_y_continuous(limits = c(0, 10), 
                       breaks = seq(0, 10, 1), 
                       labels = comma)
```
<br>
Finally, report the mean and median number of steps taken per day:  
- The mean is **`r format(round(mean(daily$numSteps)), big.mark = ",")`**  
- The median is **`r format(median(daily$numSteps), big.mark = ",")`**  

```{r}
mean(daily$numSteps)
median(daily$numSteps)
```
<br>


## What is the average daily activity pattern?
First, average the number of steps taken to the five minute interval level
```{r}
interval <- activity %>% 
    group_by(minutes, time) %>%  
    summarise(avgSteps = mean(steps, na.rm = TRUE)) %>% 
    arrange(minutes, time)
```
Next, create a time series plot of the average steps taken by time interval
```{r, fig.width=10, fig.height=6}
ggplot(data = interval, 
       aes(x = minutes, 
           y = avgSteps)) + 
    geom_line(col = "black") + 
    xlab("Time of Day (hh:mm)") + 
    ylab("Average Number of Steps") + 
    scale_x_continuous(limits = c(0, 1435), 
                       breaks = c(seq(0, 1400, 120), 1435), 
                       labels = format(as.POSIXct(c(seq(0, 1400, 120), 1435)*60, 
                                                  origin = as.Date(Sys.time()), 
                                                  tz = "UTC"), 
                                       "%H:%M")) + 
    scale_y_continuous(limits = c(0, 225), 
                       breaks = seq(0, 225, 25), 
                       labels = comma)
```
<br>
Finally, report the time interval with the maximum number of average steps
```{r, results='asis'}
## identify observation
t <- interval %>% 
    inner_join(data.frame(minutes = interval$minutes, 
                          avgRank = dense_rank(interval$avgSteps)), 
               "minutes") %>% 
    as.data.frame() %>% 
    arrange(desc(avgRank)) %>% 
    filter(row_number() == 1) %>% 
    select(minutes, time, avgSteps)

## print using xtable
suppressMessages(library(xtable))
print(xtable(t), 
      type = "html", 
      include.rownames = FALSE)
```
```{r, echo=FALSE}
rm(t)
```
<br>


## Imputing missing values
First, report the number of rows in the original data with missing values:  
- The mean is **`r format(sum(is.na(activity$steps)), big.mark = ",")`**   

```{r}
sum(is.na(activity$steps))
```
Then, using the average # of steps per interval, create a new dataset filling in the missing values
```{r}
act <- activity %>% 
    inner_join(interval, c("minutes", "time")) %>% 
    mutate(steps = ifelse(is.na(steps), 
                          avgSteps, 
                          steps)) %>% 
    select(steps, date, interval, datetime, minutes, time)
```
Next, sum the number of steps taken to the daily level
```{r}
day <- act %>% 
    group_by(date) %>% 
    summarise(numSteps = sum(steps)) %>% 
    arrange(date)
```
Now, create a new histogram of the total number of steps taken each day
```{r, fig.width=10, fig.height=6}
## histogram
ggplot(data = day, 
       aes(x = numSteps)) + 
    geom_histogram(col = "black", 
                   fill = "steelblue", 
                   binwidth = 1000) + 
    xlab("Total Number of Steps Taken Each Day") + 
    ylab("Number of Days") + 
    scale_x_continuous(limits = c(0, 25000), 
                       breaks = seq(0, 25000, 5000), 
                       labels = comma) + 
    scale_y_continuous(limits = c(0, 10), 
                       breaks = seq(0, 10, 1), 
                       labels = comma)
```
<br>
Finally, report the new mean and median number of steps taken per day:  
- The mean is **`r format(round(mean(day$numSteps)), big.mark = ",")`**  
- The median is **`r format(round(median(day$numSteps)), big.mark = ",")`**  

```{r}
mean(day$numSteps)
median(day$numSteps)
```
These new mean and median values do differ from the originals. After imputing, 
the daily totals become less right skewed and more normally distributed. The average increased by 
**`r format(round(mean(day$numSteps) - mean(daily$numSteps)), big.mark = ",")`** steps and the median increased by 
**`r format(round(median(day$numSteps) - median(daily$numSteps)), big.mark = ",")`** steps.  
<br>


## Are there differences in activity patterns between weekdays and weekends?
Using, the data with imputed values, create factor to distinguish weekdays vs. weekend days
```{r}
## create new data frame
act2 <- act %>% 
    mutate(dayType = as.factor(ifelse(as.integer(format(date, 
                                                        "%u")) %in% (1:5), 
                                      "weekday", 
                                      "weekend")))

## quickly see counts by day type
summary(act2$dayType)
```
Then, average the number of steps taken to the day type / five minute interval level
```{r}
int <- act2 %>% 
    group_by(dayType, minutes, time) %>%  
    summarise(avgSteps = mean(steps, na.rm = TRUE)) %>% 
    arrange(dayType, minutes, time)
```
Next, create a panel (faceted by day type) time series plot of the average steps taken by time interval
```{r, fig.width=10, fig.height=6}
ggplot(data = int, 
       aes(x = minutes, 
           y = avgSteps)) + 
    geom_line(col = "black") + 
    xlab("Time of Day (hh:mm)") + 
    ylab("Average Number of Steps") + 
    scale_x_continuous(limits = c(0, 1435), 
                       breaks = c(seq(0, 1400, 120), 1435), 
                       labels = format(as.POSIXct(c(seq(0, 1400, 120), 1435)*60, 
                                                  origin = as.Date(Sys.time()), 
                                                  tz = "UTC"), 
                                       "%H:%M")) + 
    scale_y_continuous(limits = c(0, 250), 
                       breaks = seq(0, 250, 50), 
                       labels = comma) + 
    facet_wrap(~ dayType, ncol = 1)
```
<br>
