---
title: 'Regression Models: Homework'
author: "Eugene Lamb"
date: "Friday, November 20, 2015"
output: html_document
---

```{r, echo=FALSE}
options(scipen=999)
```

## Count Data

### Exercises

1. Load the dataset `Seatbelts` as part of the `datasets` package via `data(Seatbelts)`. Use `as.data.frame` to convert the object to a dataframe. Fit a Poisson regression GLM with `UKDriversKilled` as the outcome and `kms`, `PetrolPrice` and `law` as predictors. Interpret your results.

```{r}
data(Seatbelts)
sb <- as.data.frame(Seatbelts)
fit <- glm(DriversKilled ~ kms + PetrolPrice + law,
           data = sb,
           family = "poisson")
summary(fit)

E <- function(x) {
    ifelse(x > 0, exp(x) - 1, -exp(-x) + 1)
}
E(coef(fit))
```

2. Refer to question 1. Fit a linear model with the log of drivers killed as the outcome. Interpret your results.

3. Refer to question 1. Fit your Poisson log-linear model with drivers as a log offset (to consider the proportion of drivers killed of those killed or seriously injured.)

4. Refer to Question 1. Use the anova function to compare models with just law, law and PetrolPrice and all three predictors.
