---
title: 'Regression Models: Quiz #4'
author: "Eugene Lamb"
date: "Saturday, April 25, 2015"
output: html_document
---

```{r, echo=FALSE}
options(scipen=999)
```

## Question 1

Consider the space shuttle data `?shuttle` in the `MASS` library.  Consider modeling the use of the autolander as the outcome (variable name `use`).  Fit a logistic regression model with autolander (variable auto) use (labeled as "auto" 1) versus not (0) as predicted by wind sign (variable wind).  Give the estimated odds ratio for autolander use comparing head winds, labeled as "head" in the variable headwind (numerator) to tail winds (denominator).


```{r, message = FALSE}
require(MASS)

data(shuttle)

fit <- glm(use ~ wind,
           data = shuttle,
           family = 'binomial')

ans <- round(exp(coef(fit))[2], 3)
```


**Answer:** `r ans`

## Question 2

Consider the previous problem.  Give the estimated odds ratio for autolander use comparing head winds (numerator) to tail winds (denominator) adjusting for wind strength from the variable `magn`.


```{r, message = FALSE}
fit <- glm(use ~ wind + magn,
           data = shuttle,
           family = 'binomial')

ans <- round(exp(coef(fit))[2], 3)
```

**Answer:** `r ans`

## Question 3

If you fit a logistic regression model to a binary variable, for example use of the autolander, then fit a logistic regression model for one minus the outcome (not using the autolander) what happens to the coefficients?

**Answer:** The coefficients reverse their signs.

## Question 4

Consider the insect spray data `InsectSprays`.  Fit a Poisson model using spray as a factor level.  Report the estimated relative rate comparing spray A (numerator) to spray B (denominator).

```{r, message = FALSE}
data(InsectSprays)
fit <- glm(count ~ factor(spray) - 1,
           data = InsectSprays,
           family = 'poisson')
X <- coef(fit)

## The expected relative increase in a unit change for regressor beta_1 = exp(beta_1)
ans <- round(exp(X)[1]/exp(X)[2], 4)
```

**Answer:** `r ans`

## Question 5

Consider a Poisson glm with an offset, ***t***. So, for example, a model of the form `glm(count ~ x + offset(t), family = 'poisson')` where `x` is a factor variable comparing a treatment (1) to a control (0) and `t` is the natural log of a monitoring time. What is impact of the coefficient for `x` if we fit the model `glm(count ~ x + offset(t2), family = 'poisson')` where `t2 <- log(10) + t`? In other words, what happens to the coefficients if we change the units of the offset variable. (Note, adding log(10) on the log scale is multiplying by 10 on the original scale.)

**Answer:** The coeficient estimate is unchanged.

## Question 6

Consider the data

```{r}
x <- -5:5
y <- c(5.12, 3.93, 2.67, 1.87, 0.52, 0.08, 0.93, 2.05, 2.54, 3.87, 4.97)
```

Using a knot point at 0, fit a linear model that looks like a hockey stick with two lines meeting at x=0. Include an intercept term, x and the knot point term. What is the estimated slope of the line after 0?

```{r}
## the break points of the spline fit
knots <- c(0)

## building the regression spline terms
splineTerms <- sapply(knots, function(knot) (x > knot) * (x - knot))

## adding an intercept and the linear term
xMat <- cbind(1, x, splineTerms)

## fit the model, notice the intercept is in xMat so we have -1
yhat <- predict(lm(y ~ xMat - 1))

## Slope = (y1 - y2) / (x1 - x2)
ans <- diff(yhat[6:7])
```

Answer: `r ans`