---
title: "Activity Monitoring"
output: html_document
---

## Part 1
Load and preprocess the data.

```{r}
unzip('activity.zip')
activity <- read.csv('activity.csv')
activity$date <- as.Date(activity$date)
activity_non_null <- subset(activity, !is.na(activity$steps))
```

## Part 2
Create a histogram of the total number of steps taken each day:

```{r, message=FALSE, echo=FALSE}
library(sqldf)
step_days <- sqldf('SELECT DATE, SUM(STEPS) AS TOTAL_STEPS 
                        FROM activity_non_null
                        GROUP BY DATE')
plot(x=step_days$date, 
     y=step_days$TOTAL_STEPS, 
     type='h', 
     lwd = 3, 
     main = 'Steps Taken Per Day', 
     xlab = 'Day', 
     ylab = 'Number of Steps', 
     col = 'red')
```

What is the mean and median total number of steps taken per day?

The Arithmetic mean is equal to 
$$ \frac{1}{n} \sum_{i=1}^{n} x_{i} $$

```{r, message=FALSE, echo=FALSE}
library(plyr)
mean_med_steps <- ddply(activity_non_null, 
                        c('date'), 
                        summarize, 
                        mean = mean(steps),
                        median = median(steps))
mean_med_steps
```

## Part 3
```{r, echo=FALSE}
avg_step_intervals <- sqldf('SELECT INTERVAL, AVG(STEPS) AS AVG_STEPS
                                 FROM activity_non_null
                                 GROUP BY INTERVAL')
plot(x=avg_step_intervals$interval, 
     y=avg_step_intervals$AVG_STEPS, 
     type='l', 
     lwd = 3, 
     main = 'Average Number of Steps Per Day (in 5-minute intervals)',,
     xlab = 'Time', 
     ylab = 'Avg. Number of Steps', 
     col = 'blue')
```

Which 5-minute interval, on average, contains the maximum number of steps?
```{r, echo=FALSE}
max_interval <- avg_step_intervals[order(-avg_step_intervals[['AVG_STEPS']]),]$interval[[1]]
```
Max Interval: `r max_interval`

## Part 4
Replace all step values in entries with "NA"" step values with the average step value for that entry's interval.  Generate the histograms and mean/median values as in parts 2 and 3 using this new data.  Compare: 
```{r, echo=FALSE}
replace_na <- function(step, interval) {
    ret <- step
    if (is.na(ret)) {
        ret <- avg_step_intervals[avg_step_intervals$interval == interval,]$AVG_STEPS
    }
    ret
}
na_count <- nrow(subset(activity, !is.na(activity$steps)))
new_activity <- activity
new_activity$steps <- mapply(replace_na,
                             new_activity$step,
                             new_activity$interval)
new_step_days <- sqldf('SELECT DATE, SUM(STEPS) AS TOTAL_STEPS 
                           FROM new_activity
                           GROUP BY DATE')
plot(x=new_step_days$date, 
     y=new_step_days$TOTAL_STEPS, 
     type='h', 
     lwd = 3, 
     main = 'Steps Taken Per Day (NAs filled)', 
     xlab = 'Day', 
     ylab = 'Number of Steps', 
     col = 'orange')
mean_med_new <- ddply(new_activity, 
                      c('date'), 
                      summarize, 
                      mean = mean(steps),
                      median = median(steps))
mean_med_new
```

With a few exceptions, replacing "NA" values increased both the total steps and mean/median number of steps per day.

## Part 5

Identify "weekday" and "weekend" entries.  Compare the daily average number of steps (per 5-minute interval) for weekdays vs. weekends.

```{r, message = FALSE, echo = FALSE}
library(ggplot2)

day_class <- function(date) {
    ret <- 'weekday'
    day <- weekdays(as.Date(date))
    if (day %in% c('Saturday','Sunday')) {
        ret <- 'weekend'
    }
    ret
}

new_activity$day.type <- sapply(new_activity$date, day_class)
avg_steps <- ddply(new_activity,
                   c('day.type','interval'),
                   summarize,
                   mean = mean(steps))


p = qplot(x = interval, 
          y = mean, 
          data = avg_steps, 
          facets = day.type ~ ., 
          geom = 'line', 
          stat = 'identity',
          main = "Daily Average Steps (per 5-minute interval)",
          xlab = '5-minute Interval',
          ylab = 'Average Steps')
print(p)
```