---
title: 'Statistical Inference Part 2: Tooth Growth'
output: pdf_document
---

# Introduction

For part 2 of our investigations into statistical inference, we shall examine the `ToothGrowth` data set.  This data set contains tooth length measurements for guinea pigs given vitamin C supplements. 

First, we will perform some basic exploratory analysis.  Then, we will provide a summary of the data and use confidence intervals to compare tooth growth by supplementation (ascorbic acid vs. orange juice) and dosage (0.5, 1.0, and 2.0 mg).  Our conclusions will be stated at the end. 

# Exploratory Analysis of ToothGrowth

Let's start by loading our data set and the necessary libraries.

```{r prelim, message=FALSE, warning=FALSE, results='hide'}
library(ggplot2) # For plots
library(scales)  # Also for plots
library(plyr)    # For data munging

options(scipen=999) # To disable scientific notation output

# Load ToothGrowth
data(ToothGrowth)
```

Now let's provide a plot for some exploratory analysis of the data.

```{r explore, warning=FALSE}
plot_explore <- function(data, 
                         title = 'The Effect of Vitamin C on Tooth Growth in Guinea Pigs',
                         subtitle = 'Grouped by Supplement Type and Dosage (in mg)'
                         ) {
    
    p <- ggplot(data, aes(x=dose, y=len, fill=factor(dose))) +
        
         # Facet by supplement
         facet_grid(. ~ supp) +
        
         # Make this plot a histogram
         geom_boxplot() +
        
         # Label our plot
         ggtitle( bquote( atop(.(title), atop(.(subtitle), '') ) ) ) +
         xlab('Dosage (in milligrams)') +
         ylab('Tooth Length') +
         
         # Change the legend label and Dosage labels
         scale_fill_discrete(name='Dosage', 
                             breaks=c(0.5,1.0,2.0), 
                             labels=c('0.5 mg', '1.0 mg', '2.0 mg'))
    p
}

plot_explore(ToothGrowth)
```

# Summary of ToothGrowth

Here we will provide an overall summary of the data in `ToothGrowth`

```{r summary, warning=FALSE}
summary(ToothGrowth)
```

# Confidence Testing

Now let us investigate the means and confidence intervals (95%) of tooth length by supplement type and dosage.  Recall that the 95% confidence interval for a sample mean is:

$$
    \mu \pm 2\sigma/\sqrt{n}     
$$

By supplement:

```{r test_supp, warning=FALSE}
by_supp <- ddply(ToothGrowth, 
                 .(supp),
                 summarize,
                 mean=mean(len), 
                 conf.int=2 * sd(len) / sqrt( length(len) ))
by_supp$lower.bound <- with(by_supp, mean - conf.int)
by_supp$upper.bound <- with(by_supp, mean + conf.int)
by_supp
```

By dosage:

```{r test_dose, warning=FALSE}
by_dose <- ddply(ToothGrowth, 
                 .(dose),
                 summarize,
                 mean=mean(len), 
                 conf.int=2 * sd(len) / sqrt( length(len) ))
by_dose$lower.bound <- with(by_dose, mean - conf.int)
by_dose$upper.bound <- with(by_dose, mean + conf.int)
by_dose
```

# Conclusions

There is a positive correlation between the dosage of vitamin C and tooth growth for guinea pigs.  The 2.0 mg dosage provided the greatest tooth growth, with a mean of $26.1 \pm 1.69$.  This interval is safely above the intervals for 1.0 mg and 0.5 mg with a 95% confidence.

It also appeared that ascorbic acid (mean $20.66 \pm 2.41$) had a greater effect than orange juice (mean $16.96 \pm 3.02$) for tooth growth, but since their confidence intervals overlapped, more data is needed to determine the true benefits of different supplementation types.
